\chapter{Versioning and support policy}
The PostgreSQL version's number is composed by three integer. The first number is the generational
version. Currently the value is 9. This number changes when there is a substantial generational gap
with the previous version. For example the version 9.0 started its life as 8.5. Later was decided
the change of generation.\newline

The second number is the major version. The value starts from zero and increases by one for each
release along the generation. Because each major version differs internally from the others the data
area is not compatible between them. Usually a new major version is released on a yearly
basis.\newline

The third number is the minor version. Usually a new minor release appears when a sufficient number
of bug fixes are merged into the codebase. Upgrading a minor version usually requires just the
binary upgrade and the cluster restart. However is a good practice to check for any extra action
required in the release notes.\newline

The PostgreSQL project aims to fully support a major release for five years.  The policy is
applied on a best-effort basis. 

\begin{table}[H]
  \begin{tabular}{cccc}
    Version & Supported & First release date & End of life date \\ 
    \hline
    9.3 & Yes &September 2013 & September 2018\\
    9.2 & Yes & September 2012 & September 2017\\
    9.1 & Yes & September 2011 & September 2016\\
    9.0 & Yes & September 2010 & September 2015\\
    8.4 & No & July 2009 & July 2014\\
    
  \end{tabular}
  \caption{\label{tab:EOLDATES}End Of Life (EOL) dates}
\end{table}