\chapter{Install structure}\label{cha:INSTALLSTRUCT}
Depending on the installation method, the install structure is set up in a single directory or 
in multiple folders.\newline

The install from source creates four subfolders in the target directory: \textit{bin},
\textit{include}, \textit{lib} and \textit{share}.

\begin{itemize}
 \item \textbf{bin} contains the PostgreSQL binaries
 \item \textbf{include} contains the server's header files
 \item \textbf{lib} contains the shared libraries
 \item \textbf{share} contains the example files and the extension configurations
\end{itemize}


The packaged install puts the binaries and the libraries in the folder /usr/lib/postgresql 
organised by major version. For example the 9.3 install will put the binaries into 
/usr/lib/postgresql/9.3/bin and the libraries in /usr/lib/postgresql/9.3/lib. The extensions and 
contributed modules are installed into the folder /usr/share/postgresql with the same structure. The
Debian specific utilities and the symbolic link to the psql binary (which is at 
/usr/lib/share/postgresql-common/pg\_wrapper) are stored in the directory /usr/bin/. This file is a perl script which 
calls the PostgreSQL client reading the version the cluster and the default database from the file 
~/.postgresqlrc or /etc/postgresql-common/user\_clusters.\newline


\section{The core binaries}
The PostgreSQL binaries can be split in two groups, the core and the wrappers alongside with the 
contributed modules. Let's start then with the former group.

\subsection{postgres}\index{postgres}
\label{sec:POSTGRES}
This is the PostgreSQL's main process. The program can be started directly or using the pg\_ctl 
utility. The second method is to be preferred as it offers a simpler way to control the postgres 
process. Direct execution is the unavoidable choice when the database won't start for an old XID 
near to the wraparound failure\index{XID wraparound failure}. 
In this case the cluster can only start in single user mode to perform a cluster wide vacuum. For 
historical reasons there's also a symbolic link named postmaster pointing to the postgres 
executable.

\subsection{pg\_ctl}\index{pg\_ctl}
\label{sub:PGCTL}
This utility is the simplest way for managing a PostgreSQL instance. The program reads the postgres 
pid from the cluster's data area and sends the os signals to start, stop or reload the
process. It's also possible to send kill signals to the running instance. 
pg\_ctl provides the following actions:

\begin{itemize}
 \item \textbf{init[db]} initialises a directory as PostgreSQL data area
 \item \textbf{start} starts a PostgreSQL instance
 \item \textbf{stop} shutdowns a PostgreSQL instance
 \item \textbf{reload} reloads the configuration's files
 \item \textbf{status} checks the PostgreSQL instance running status
 \item \textbf{promote} promotes a standby server 
 \item \textbf{kill} sends a custom signal to the running instance
\end{itemize}

In \ref{cha:MANAGING} we'll se how to manage the cluster.

\subsection{initdb}\index{initdb}
Is the binary which initialises the PostgreSQL data area. The directory to initialise must 
be empty. Various options can be specified on the command line, like the character encoding or the 
collation order. 

\subsection{psql}\index{psql}
% TODO: Review statement 'The client it looks very essential'
This is the PostgreSQL command line client. The client it looks very essential, however is one of 
the most flexible tools available to interact with the server and the only choice when working on 
the command line.

\subsection{pg\_dump}\index{pg\_dump}
\label{sub:PGDUMP}
This is the binary dedicated to backup. Can produce consistent backups in various formats. The 
usage is described shown in \ref{cha:BACKUP}.

\subsection{pg\_restore}\index{pg\_restore}
This program is used to restore a database reading a binary dump like the custom or directory 
format. It's able to run the restore in multiple jobs in order to speed up the process. The usage 
is described in \ref{cha:RESTORE}

\subsection{pg\_controldata}\index{pg\_controldata}\label{sub:PGCONTROLDATA}
This program can query the cluster's control file where PostgreSQL stores critical information about 
the cluster activity and reliability. 

\subsection{pg\_resetxlog}\index{pg\_resetxlog}
If a WAL file becomes corrupted the cluster cannot perform a crash recovery. This lead to an
unstartable cluster in case of system crash. In this catastrophic scenario there's still a 
way to start the cluster. Using pg\_resetxlog the cluster is cleared of any WAL file, the  
control file is initialised from scratch and the transaction's count is restarted.\newline

The \textit{tabula rasa} comes with a cost indeed. The cluster loses any reference between the 
transaction progression and the data files. The physical integrity is lost and any attempt to run 
queries which write data will result in corruption.\newline 

The PostgreSQL's documentation is absolutely clear on this point.

\begin{verbatim}

After running pg_resetxlog the database must start without user access, 
the entire content must be dumped, the data directory must be dropped and recreated 
from scratch using initdb and then the dump file can be restored using psql or pg_restore
\end{verbatim}

\section{Wrappers and contributed modules}
The second group of binaries is composed of the contributions and the wrappers. The 
contributed modules add functions otherwise not available. The wrappers add command line 
functions already present as SQL statements. Someone will notice the lack of HA specific binaries 
like pg\_receivexlog and pg\_archivecleanup. They have been purposely skipped because they are beyond the 
scope of this book.

\subsection{The create and drop utilities}
The binaries with the prefix create and drop like, createdb createlang createuser and dropdb, 
droplang, dropuser, are wrappers for the corresponding SQL functions. Each program performs the 
creation and the drop action on the corresponding named object. For example createdb adds a 
database to the cluster and dropdb will drop the specified database. 

\subsection{clusterdb}\index{clusterdb}
This program performs a database wide cluster on the tables with clustered indices. 
The binary can run on a single table specified on the command line. In \ref{sec:VACFULL} we'll 
take a look to CLUSTER and VACUUM FULL.

\subsection{reindexdb}\index{reindexdb}
The command does a database wide reindex. It's possible to run the command just on a table or index 
passing the relation's name on the command line. In \ref{sec:REINDEX} we'll take a good look at 
the index management.

\subsection{vacuumdb}\index{vacuumdb}
This binary is a wrapper for the VACUUM \index{VACUUM} SQL command. This is the most important 
maintenance task and shouldn't be ignored. The program performs a database wide VACUUM if executed 
without a target relation. Alongside with a common vacuum it's possible to have the usage 
statistics updated on the same time.

\subsection{vacuumlo}\index{vacuumlo}
This binary will remove the orphaned large objects from the pg\_largeobject system table. The 
pg\_largeobject is used to store the binary objects bigger than the limit of 1GB imposed by the 
bytea data type. The limit for a large object it is 2 GB since the version 9.2. From the version 
9.3 the limit was increased to 4 TB. 

\section{Debian's specific utilities}
Finally let's take a look at the Debian specific utilities. They are a collection of perl scripts 
used to simplify the cluster's management. They are installed in /usr/bin and mostly consist of symbolic 
links to the actual executable. We already mentioned one of them in the chapter's introduction, the 
psql pointing to the pg\_wrapper PERL script.

\subsection{pg\_createcluster}\index{pg\_createcluster}
This script adds a new PostgreSQL cluster with the given major version, if installed, and the 
given name. The script puts all the configuration in /etc/postgresql. Each major version has a 
dedicated directory within which is a group of directories containing the cluster's specific 
configuration files. If not specified the data directory is created in the folder 
/var/lib/postgresql. It's possible to specify the options for initd.

\subsection{pg\_dropcluster}\index{pg\_dropcluster}
The program will delete a PostgreSQL cluster created previously with pg\_createcluster. The 
program will not drop a running cluster. If the dropped cluster has any tablespace those must be 
manually removed after the drop as the program doesn't follow the symbolic links.

\subsection{pg\_lscluster}\index{pg\_lscluster}
Lists the clusters created with pg\_createcluster.

\subsection{pg\_ctlcluster}\index{pg\_ctlcluster}
\label{sub:PGCTLDEB}
% TODO: Understand this better to rewrite
The program manages the cluster in a similar way pg\_ctl does. 
Before version 9.2 this wrapper had dangerous behaviour for the shutdown. The script did not 
offered a flexible way to provide the shutdown mode. More informations about the shutdown 
sequence are in \ref{sec:SHUTDOWN_SEQ}. 
When run without any arguments pg\_ctlcluster performs a smart shutdown mode.
The --force option tells the script to try a \textit{fast} shutdown mode. Unfortunately if the 
database doesn't shutdown in a \textit{reasonable time} the script performs an \textit{immediate} 
shutdown. After another short wait, if the the instance is still up the script sends a 
\textit{kill -9} to the postgres process. Because these kind of actions can result in data loss  
they should be done manually by the DBA. It's better to avoid using pg\_ctlcluster for the shutdown.
